\documentclass{article}
\usepackage{a4}
\setlength{\oddsidemargin}{20pt}
\addtolength{\textwidth}{39pt}
\newcommand{\var}[1]{{\texttt #1}}
\usepackage{tabularx}
\begin{document}
\thispagestyle{empty}
%\section*{Compiler options and command-line switches}
\subsection*{Local compiler switches}
%\begin{table}
%\caption[b]{Local compiler switches.}
\begin{tabularx}{\textwidth}{lllX}
\textbf{cmd} & \textbf {short} & \textbf {long} & \textbf{explanation }\\ \hline
& \var{\$A} & \var{\$ALIGN} & Align Data.\\
A & & \var{\$ASMMODE} & Select assembler mode. \\
&\var{\$B} & \var{\$BOOLEVAL} & Use complete boolean evaluation. \\
Sa &\var{\$C} & \var{\$ASSERTIONS} & Enable assertion support. \\
& & \var{\$CALLING} & Set default calling convention \\
d && \var{\$DEFINE} & Define a symbol. \\
&& \var{\$ELSE} & Switch conditional compilation. \\
&& \var{\$ENDIF} & End conditional compilation. \\
&& \var{\$ERROR} & Generate error message. \\
&\var{\$F} & & Use far or near functions. \\
&&  \var{\$FATAL} & Generate fatal error message. \\
Sg && \var{\$GOTO} & Support \var{Goto} and \var{Label}. \\ 
&\var{\$H} & \var{\$LONGSTRINGS} & Use AnsiStrings. \\
 && \var{\$HINT} & Generate hint message. \\
vh && \var{\$HINTS} & Emit hints \\
&& \var{\$IF} & Start conditional compilation. \\
&& \var{\$IFDEF} & Start conditional compilation. \\
&& \var{\$IFNDEF} & Start conditional compilation. \\
&& \var{\$IFOPT} & Start conditional compilation. \\
&& \var{\$INFO} & Generate info message. \\
Si && \var{\$INLINE} & Enable inline code support. \\
Ci &\var{\$I} & \var{\$IOCHECKS} & Include Input/Output checking. \\
&\var{\$I} & \var{\$INCLUDE} & Include file.  \\
&\var{\$I} & \var{\$INCLUDE} & Include compiler info. \\
&\var{\$L} & \var{\$LINK} & Link object file. \\
&& \var{\$LINKLIB} & Link to a library. \\
&\var{\$M} & \var{\$TYPEINFO} & Generate Run-Time type information. \\
Sm & & \var{\$MACRO} & Enable macro support. \\
& & \var{\$MAXFPUREGISTER} & set maximum number of FPU registers \\
&& \var{\$MESSAGE} & Generate info message. \\
&& \var{\$MMX} & Enable Intel MMX support. \\
&& \var{\$NOTE} & Generate note message. \\
vn && \var{\$NOTES} & Emit notes. \\
A && \var{\$OUTPUT\_FORMAT} & Select compiler output format. \\
&\var{\$P} & \var{\$OPENSTRINGS} & Use open strings. \\
&& \var{\$PACKENUM} & Specify minimum enumeration type size. \\
&& \var{\$PACKRECORDS} & Specify Alignment of record elements. \\
Co &\var{\$Q} & \var{\$OVERFLOWCHECKS}& Use overflow checking. \\
Cr &\var{\$R} & \var{\$RANGECHECKS} & Use range checking. \\
&& \var{\$SATURATION} & Enable saturation operations. \\
XX && \var{\$SMARTLINK} & Use smartlinking. \\
St && \var{\$STATIC} & Enable use of \var{Static} keyword. \\
&& \var{\$STOP} & Generate fatal error message. \\
&\var{\$T} & \var{\$TYPEDADDRESS} & Enable typed address operator. \\
u && \var{\$UNDEF} & Undefine a symbol. \\
&\var{\$V} & \var{\$VARSTRINGCHECKS} & Use strict var-string checking. \\
&& \var{\$WAIT} & Wait for enter key press. \\
&& \var{\$WARNING} & Generate warning message. \\
&& \var{\$WARNINGS} & Emit warnings. \\
&\var{\$X} & \var{\$EXTENDEDSYNTAX} & Enable use of extended syntax. \\ \hline
\end{tabularx}
%\end{table}
\subsection*{Global compiler switches}
%\begin{table}
%\caption{Global compiler switches}
\begin{tabularx}{\textwidth}{lllX}
\textbf{cmd} & \textbf {short} & \textbf {long} & \textbf{explanation }\\ \hline
& & \var{\$APPID} & Set application ID (PalmOS) \\
& & \var{\$APPNAME} & Set application name (PalmOS) \\
& & \var{\$APPTYPE} & Specify type of application (Win32 only) \\
g& \var{\$D} & \var{\$DEBUGINFO} & Include debugging symbols. \\
& & \var{\$DESCRIPTION} & Not supported. \\
&\var{\$E} & & Enable emulation of coprocessor. \\
& \var{\$G}& & Generate 80286 code. \\
Fi& & \var{\$INCLUDEPATH} & Specify include file search path. \\
&\var{\$L} & \var{\$LOCALSYMBOLS} & Enable local symbol information. \\
Fl& & \var{\$LIBRARYPATH} & Specify library search path. \\
&\var{\$M} & \var{\$MEMORY} & Specify memory sizes. \\
M& & \var{\$MODE} & Specify compiler compatibility mode. \\
& \var{\$N}& & Enable numeric processing.  \\
& \var{\$O}& & Enable overlay code generation.  \\
Fo& & \var{\$OBJECTPATH} & Specify object file search path. \\
Ct& \var{\$S} & & Use stack checking \\
Fu& & \var{\$UNITPATH} & Specify unit file search path. \\
& & \var{\$VERSION} & Set DLL version (Windows)\\
& \var{\$W} & \var{\$STACKFRAMES} & Generate stackframes. \\
b & \var{\$Y} & \var{\$REFERENCEINFO} & Insert browser information. \\ \hline
\end{tabularx}
%\end{table}
\end{document}